---
title: "Okanagan Chinook Population Viability Analysis"
output: html_notebook
editor_options: 
  chunk_output_type: inline
---

##Background


## Premable

Before this analysis can be run you will need the following packages installed
in your R distribution.
```{r}
package.list <- c('devtools', 'tidyverse', 'ggplot2', 'scales', 'gridExtra', 'openxlsx')
```

Next we will installs any packages that may be missing.
```{r, echo=F}
## Attempt to install missing packages. 
new.packages <- package.list[!(package.list %in% installed.packages()[,"Package"])]
if(length(new.packages) > 0) {
  install.packages(new.packages)
} else {
  message("All required packages have already been installed.")
}
```

With the packages installed we can load the custom CNPVA r-package that 
is used to do the calculations along with any other required packages.
```{r}
rm(list=ls())  # clear workspace
library(devtools)  # Sideload CNPVA custom R package
library(openxlsx) # Useful for creating excel files
load_all("CNPVA")
library(grid)
library(gridExtra)
theme_set(theme_bw(14))
```

Finally, we will create an excel workbook to save our analysis output into.

```{r}
wb <- createWorkbook()
wb.out <- createWorkbook()
```




## Settings

First we create a list object to save the simulation scenario results.
```{r}
# nsim <- 1000     # Development
nsim <- 10000    # Final run

all.runs <- list()
results <- list(
  plots = list(spawners = list(), adults = list()), 
  data = list(spawners=list(), adults=list())
)
```

### Default Settings
Currently the S.osar represent survival from MCN dam to age 2.
```{r}
setdef <- GetSettings("default v2")
str(setdef)
```

Compute expected returns for the 15,000 releases
```{r}
n.rel <- 15000

S.eo <- setdef$S.osar
HR.o <- setdef$hr.ocean
S.o  <- setdef$surv.ocean
Spwn <- setdef$spawn.prob
S.r  <- setdef$S.riv
HR.r <- setdef$hr.river

spawners <- rep(NA,4)
names(spawners) <- paste0("age.", 2:5)

spawners['age.2'] <- n.rel * S.eo * (1 - HR.o) *  Spwn[1] * S.r * (1 - HR.r)       # Age 2 SPawner
spawners['age.3'] <- n.rel * S.eo * (1 - HR.o) * (1 - Spwn[1]) * S.o[2] * (1 - HR.o) *  Spwn[2] * S.r * (1 - HR.r)    # Age 3 Spawner
spawners['age.4'] <- n.rel * S.eo * (1 - HR.o) * (1 - Spwn[1]) * S.o[2] * (1 - HR.o) * (1 - Spwn[2]) * S.o[3] * (1 - HR.o) * Spwn[3] * S.r * (1 - HR.r) # Age 4 Spawner 
spawners['age.5'] <- n.rel * S.eo * (1 - HR.o) * (1 - Spwn[1]) * S.o[2] * (1 - HR.o) * (1 - Spwn[2]) * S.o[3] * (1 - HR.o) * (1 - Spwn[3]) * S.o[4] * (1 - HR.o) * 1 * S.r * (1 - HR.r) 

spawners
```

```{r}
round(sum(spawners))
```



### Upriver survival

Load the upstream survival data included in the CNPVA package
```{r}
data(upstream)
upstream
```
```{r, echo=FALSE}
ggplot(upstream, aes(x=Year, y=Survival)) + 
  geom_point() +
  scale_y_continuous(trans=logit_trans(), breaks=c(seq(0.2, 0.8, by=0.2), 0.95, 0.99))  + 
  geom_smooth(method="lm", formula = y~poly(x,1), se = TRUE) +
  ylab("Upstream Survival")
```

Next we will create a linear model fit object containing the above regression
relationship:
```{r}
summary(fit <- lm(qlogis(Survival)~Year, upstream))
```

This empirical regression relationship is used in the PVA to predict future
survival rates with uncertainty (i.e., prediction intervals). To do so we must
include the fit object in the settings passed to the simulator
```{r}
setdef$S.riv <- fit
```




## Scenario Runs

The PVA considered a number of future scenarios over which to assess population
viability these included:

1. Mortality reductions through habitat improvements
2. Hatchery supplementation
3. Additional scenarios which combined aspects of (1) and (2)

In this section we run the individual scenarios and add each run as a list 
element in the `all.runs` object.

However, first we must run the baseline or "status quo" scenario which captures
our expectations of the future if no changes are implemented.

```{r}
all.runs[['Baseline']] <- PopSim(setdef, n.sim=nsim)
```


### Mortality Reductions (Habitat Improvements)

It is too early in the recovery process to discuss what survival improvements
associated with a particular habitat improvement strategy.  As such, we considered
reductions in juvenile mortality, which we apply to the `S.juv` parameter.

#### Scenarios

A 10% reduction in juvenile mortality.
```{r}
settings <- setdef
settings$S.juv <-  1 - ((1- settings$S.juv) * (1 - 0.10))
all.runs[['10% mort reduction']] <- PopSim(settings, n.sim = nsim)
```


A 30% reduction in juvenile mortality.
```{r}
settings <- setdef
settings$S.juv <-  1 - ((1- settings$S.juv) * (1 - 0.30))
all.runs[['30% mort reduction']] <- PopSim(settings, n.sim = nsim)
```


A 50% reduction in juvenile mortality.
```{r}
settings <- setdef
settings$S.juv <-  1 - ((1- settings$S.juv) * (1 - 0.5))
all.runs[['50% mort reduction']] <- PopSim(settings, n.sim = nsim)
```

#### Results

```{r, echo=FALSE}
# Process results and save to excel
runs <- c(
  "Baseline", 
  "10% mort reduction", 
  "30% mort reduction", 
  "50% mort reduction"
)
run.set <- all.runs[runs]


# Compute summary tables
results[['data']][['spawners']][['Habitat']] <- spawn.tab <- SummarizeSpawners(run.set, start.yr=2022)
results[['data']][['adults']][['Habitat']] <- adults.tab <- SummarizeMetrics(run.set)    # Adult performance metrics

# library(openxlsx)
# wb <- createWorkbook()
BlankWorkSheet(wb, sheet = "Habitat" )
writeData(wb, "Habitat", select(spawn.tab$All, Type:SD))
writeData(wb, "Habitat", select(adults.tab$All, Type:SD), startCol = 6)
```

#### Outcome Table

```{r}
(out1.tab <- OutcomeTable(run.set))
BlankWorkSheet(wb.out, sheet = "Habitat")
writeData(wb.out, "Habitat", out1.tab)
```




#### Overview

```{r, fig.width=8, fig.height=6,  echo=FALSE}

p.spawn <- ggplot(spawn.tab$All, aes(x=Scenario, y=Median)) + 
  # geom_bar(stat="identity", fill="grey85", color="black") + 
  geom_point() +
  geom_errorbar(aes(ymin=Lower1, ymax=Upper99), width=0.25) +
  facet_wrap(~Type,scales = "free_x") +
  scale_y_continuous(labels = comma) +
  # geom_hline(yintercept = 1000, linetype = "dashed")+
  geom_hline(data = data.frame(Type =c("Average Spawners"), y = c( 1000)), mapping = aes( yintercept = y), linetype = "dashed") +
  geom_hline(data = data.frame(Type =c("Minimum Spawners", "Average Spawners"), y = c(3500, 3500)), mapping = aes( yintercept = y), linetype = "solid", size=0) +
  coord_flip() +
  theme(
    # axis.text.x = element_text(angle=-45, hjust=0, vjust=0.5)
    axis.title.y = element_blank()
  ) +
  ylab("Number of Spawners")

# print(p.spawn)

results[['plots']][['spawners']][['Habitat']] <-  p.spawn 


p.adult <- ggplot(adults.tab$All, aes(x=Scenario, y=Mean)) + 
  # geom_bar(stat="identity", fill="grey85", color="black") + 
  geom_point() +
  geom_errorbar(aes(ymin=Lower95, ymax=Upper95), width=0.25) +
  facet_wrap(~Type, scales = "free_x") +
  # scale_y_continuous(labels = comma, breaks = c(seq(0,10000, by=2500))) +
   scale_y_continuous(labels = comma) +
  # geom_hline(data = data.frame(Type =c("Minimum Population Size", "Population in 2050"), y = c(2500, 5000)), mapping = aes( yintercept = y), linetype = "dashed") +
  geom_hline(data = data.frame(Type =c("Minimum Population Size", "Population in 2050"), y = c(7500, 15000)), mapping = aes( yintercept = y), linetype = "solid", size=0) +
  # geom_hline()
  coord_flip() +
  theme(
    # axis.text.x = element_text(angle=-45, hjust=0, vjust=0.5)
    axis.title.y = element_blank()
  ) +
  ylab("Number of Adults")

results[['plots']][['adults']][['Habitat']] <- p.adult


grid.arrange(
  results[['plots']][['spawners']][['Habitat']] + ggtitle("A) Spawners"),
  results[['plots']][['adults']][['Habitat']] + ggtitle("B) Adults"),
  nrow=2, ncol=1,
  as.table = F,
  clip=F
)

```

#### Detailed Spawners


```{r, fig.width=10, fig.height = 5}
PlotSpawners(run.set)
```

```{r}
spawn.tab$Minimum
```
```{r}
spawn.tab$Average
```


#### Detailed Adults

```{r, fig.width=8, fig.height=8}
PlotAdultMetrics(run.set, bins=100)
```

#### Alternative Metrics

Empirial cumulative distribution plots of performance metrics.
```{r, fig.width=12, fig.height=4}
p.ecdf <- PlotMetricECDF(run.set, type="both", panel.widths = c(1,1.75))
# ggsave(filename = "habitat-ecdf.png", width=10, height=4, plot=p.ecdf)
```


```{r}
adults.tab$MPS
```

```{r}
adults.tab$POP2050
```


### Hatchery Supplementation

Hatchery supplementation looks at the effect of supplementing the wild population
with hatchery releases each year.  Smolts are assumed to be full fitness, for
reduced fitness can be consider for a given result by increasing the actual
number of smolts release by what is believed to be the fitness deficit.
#### Scenarios
Hatchery Augmentation 50K (full fitness)
```{r}
settings <- GetSettings('default v2')
settings$supplementation = 5 * 10^4
all.runs[['Hatchery 50K']] <- PopSim(settings, n.sim = nsim)
```


 Hatchery Augmentation 100K (full fitness)
```{r}
settings <- GetSettings('default v2')
settings$supplementation = 1 * 10^5
all.runs[['Hatchery 100K']] <- PopSim(settings, n.sim = nsim)
```


 Hatchery Augmentation 150K (full fitness)
```{r}
settings <- GetSettings('default v2')
settings$supplementation = 1.5 * 10^5
all.runs[['Hatchery 150K']] <- PopSim(settings, n.sim = nsim)
```


 Hatchery Augmentation 250K (full fitness)
```{r}
settings <- GetSettings('default v2')
settings$supplementation = 2.5 * 10^5
all.runs[['Hatchery 250K']] <- PopSim(settings, n.sim = nsim)
```


 Hatchery Augmentation 500K (full fitness)
```{r}
settings <- GetSettings('default v2')
settings$supplementation = 5 * 10^5
all.runs[['Hatchery 500K']] <- PopSim(settings, n.sim = nsim)
```


#### Results

```{r, echo=FALSE}
# Process results and save to excel
runs <- c(
  "Baseline", 
  "Hatchery 50K",
  "Hatchery 100K",
  "Hatchery 150K",
  "Hatchery 250K",
  "Hatchery 500K"
  )
run.set <- all.runs[runs]

# Compute summary tables
results[['data']][['spawners']][['Element20']] <- spawn.tab <- SummarizeSpawners(run.set, start.yr=2022)
results[['data']][['adults']][['Element20']] <- adults.tab <- SummarizeMetrics(run.set)    # Adult performance metrics

BlankWorkSheet(wb,sheet = "Hatchery" )
writeData(wb, "Hatchery", select(spawn.tab$Minimum, Type:SD))
writeData(wb, "Hatchery", select(spawn.tab$Average, Type:SD), startCol = 6)
```


#### Outcome Table
```{r}
(out2.tab <- OutcomeTable(run.set))
BlankWorkSheet(wb.out, sheet = "Hatchery")
writeData(wb.out, "Hatchery", out2.tab)
```


#### Overview
```{r, fig.width=8, fig.height=6, echo=FALSE}
results[['plots']][['spawners']][['Hatchery']] <-  p.spawn %+% spawn.tab$All
results[['plots']][['adults']][['Hatchery']] <- p.adult %+% adults.tab$All
grid.arrange(
  results[['plots']][['spawners']][['Hatchery']] + ggtitle("A) Spawners"),
  results[['plots']][['adults']][['Hatchery']] + ggtitle("B) Adults"),
  nrow=2, ncol=1,
  as.table = F,
  clip=F
)

```





#### Detailed Spawners

```{r}
PlotSpawners(run.set)
# ggsave(filename = "hatchery-scenarios--spawners.png", width=10, height=8, plot=p)
```
```{r}
spawn.tab$Minimum
```
```{r}
spawn.tab$Average
```




#### Detailed Adults
```{r, fig.width=10, fig.height=8, echo=FALSE}
p.met <- PlotAdultMetrics(run.set, bins=100)
```


#### Alternative metrics

```{r}
adults.tab$MPS
```

```{r}
adults.tab$POP2050
```


Empirial cumulative distribution plots of performance metrics.
```{r, fig.width=12, fig.height=4}
p.ecdf <- PlotMetricECDF(run.set, type="both", panel.widths = c(1,1.75))
# ggsave(filename = "hatchery-ecdf.png", width=10, height=4, plot=p.ecdf)
```



### Additional Scenarios

#### Scenarios

No harvest (freshwater or marine)
```{r}
settings <- GetSettings('default v2')
settings$hr.river <- 0
settings$hr.ocean <- 0
all.runs[['No Harvest']] <- PopSim(settings, n.sim = nsim)
```


A natural doubling of productivity.
```{r}
settings <- GetSettings('default v2')
settings$rec.alpha <- settings$rec.alpha *2
all.runs[['Productivity 2x']] <- PopSim(settings, n.sim = nsim)
```



150K hatchery program and a 30% reduction in juvenile mortality.
```{r}
settings <- GetSettings('default v2')
settings$supplementation = 1.5 * 10^5    # 150K
settings$S.juv <-  1 - ((1- settings$S.juv)* (1 - 0.30))
all.runs[['Hatchery 150K + 30% Mort Reduction']] <- PopSim(settings, n.sim = nsim)
```

150K hatchery program and no harvest (freshwater or marine).
```{r}
settings <- GetSettings('default v2')
settings$supplementation = 1.5 * 10^5    # 150K
settings$hr.river <- 0
settings$hr.ocean <- 0
# settings$S.juv <-  1 - ((1- settings$S.juv)* (1 - 0.30) * (1 - 0.25)* (1 - 0.1))
# settings$S.osar = settings$S.osar * 2
all.runs[['Hatchery 150K + No Harvest']] <- PopSim(settings, n.sim = nsim)
```

No harvest (freshwater or marine) and a 30% reduction in juvenile mortality.
```{r}
settings <- GetSettings('default v2')
settings$hr.river <- 0
settings$hr.ocean <- 0
settings$S.juv <-  1 - ((1- settings$S.juv)* (1 - 0.30))
# settings$S.osar = settings$S.osar * 2
all.runs[['No Harvest + 30% Mort Reduction']] <- PopSim(settings, n.sim = nsim)
```





#### Results

```{r}
runs <- c(
  "No Harvest",
  "Productivity 2x",
  # "Hatchery 250K",  
  "Hatchery 150K + 30% Mort Reduction",
  "Hatchery 150K + No Harvest",
  "No Harvest + 30% Mort Reduction"
  )
run.set <- all.runs[runs]

# Compute summary tables
results[['data']][['spawners']][['Other']] <- spawn.tab <- SummarizeSpawners(run.set, start.yr=2022)
results[['data']][['adults']][['Other']] <- adults.tab <- SummarizeMetrics(run.set)    # Adult performance metrics
```



#### Outcome Table
```{r, echo=FALSE}
(out.tab <- OutcomeTable(run.set))
BlankWorkSheet(wb.out, sheet = "Additional")
writeData(wb.out, "Additional", out.tab)
```


#### Overview
```{r, fig.width=8, fig.height=6, echo=FALSE}
results[['plots']][['spawners']][['Other']] <-  p.spawn %+% spawn.tab$All
results[['plots']][['adults']][['Other']] <- p.adult %+% adults.tab$All
grid.arrange(
  results[['plots']][['spawners']][['Other']] + ggtitle("A) Spawners"),
  results[['plots']][['adults']][['Other']] + ggtitle("B) Adults"),
  nrow=2, ncol=1,
  as.table = F,
  clip=F
)
```


#### Detailed Spawners

```{r}
PlotSpawners(run.set)
```
```{r}
spawn.tab$Minimum
```
```{r}
spawn.tab$Average
```




#### Detailed Adults
```{r, fig.width=10, fig.height=8, echo=FALSE}
p.met <- PlotAdultMetrics(run.set, bins=100)
```


#### Alternative metrics

```{r}
adults.tab$MPS
```

```{r}
adults.tab$POP2050
```


Empirial cumulative distribution plots of performance metrics.
```{r, fig.width=12, fig.height=4}
p.ecdf <- PlotMetricECDF(run.set, type="both", panel.widths = c(1,1.75))
# ggsave(filename = "hatchery-ecdf.png", width=10, height=4, plot=p.ecdf)
```


----

## Post Processing and Figures
```{r}
save(all.runs, file = "sim-runs.RData")
```


```{r}
saveWorkbook(wb, "results.xlsx", overwrite = TRUE)
saveWorkbook(wb.out, "outcomes.xlsx", overwrite = TRUE)
```


```{r}
str(all.runs, max.level = 1)
```


```{r, fig.width=9, fig.height=5.5, echo=FALSE}
library(grid)
library(gridExtra)
grid.arrange(
g <- arrangeGrob(
  results$plots$spawners$Habitat + ggtitle("A) Habitat Restoration Work") + 
    theme(
      title = element_text(size = 11),
      axis.title = element_blank(),
       axis.text.x = element_blank(),
      axis.text.y = element_text(size=11),
      plot.margin = unit(c(0.5, 0.5, 0.5, 6.75), "lines")
      ),
  results$plots$spawners$Hatchery  + ggtitle("B) Hatchery Supplementation") + 
    theme(
      title = element_text(size = 11),
      axis.title = element_blank(),
      axis.text.x = element_blank(),
      axis.text.y = element_text(size=11),
      plot.margin = unit(c(0.5, 0.5, 0.5, 8.25), "lines")
    ), 
  results$plots$spawners$Other  + ggtitle("C) Additional Scenarios") + 
    theme(
      title = element_text(size = 11),
      axis.title = element_blank(),
      axis.text.y = element_text(size=11)
    ), 
  nrow=3, ncol=1,
  heights = c(1,1,1.05),
  as.table = F,
  bottom =  textGrob( "Number of Spawners", hjust = -0.35, gp=gpar(fontsize=12, fontface="plain")),
  clip=F
))
ggsave(filename = "spawners--all-scenarios-wide.png", width=9, height=6,plot=g)
ggsave(filename = "spawners--all-scenarios-tall.png", width=9, height=8,plot=g)
```





```{r, fig.width=9, fig.height=6, echo=FALSE}

library(grid)
library(gridExtra)
grid.arrange(
g <- arrangeGrob(
  results$plots$adults$Habitat + ggtitle("A) Habitat Restoration Work") + 
    theme(
      title = element_text(size = 11),
      axis.title = element_blank(),
      axis.text.x = element_blank(),
      axis.text.y = element_text(size=11),
      plot.margin = unit(c(0.5, 0.5, 0.5, 6.75), "lines")
      ),
  results$plots$adults$Hatchery  + ggtitle("B) Hatchery Supplementation") + 
    theme(
      title = element_text(size = 11),
      axis.title = element_blank(),
      axis.text.x = element_blank(),
      axis.text.y = element_text(size=11),
      plot.margin = unit(c(0.5, 0.5, 0.5, 8.25), "lines")
    ), 
  results$plots$adults$Other  + ggtitle("C) Additional Scenarios") + 
    theme(
      title = element_text(size = 11),
      axis.title = element_blank(),
      axis.text.y = element_text(size=11)
    ), 
  nrow=3, ncol=1,
  as.table = F,
  heights = c(1,1,1.05),
  bottom =  textGrob( "Number of Adults", hjust = -0.45, gp=gpar(fontsize=12, fontface="plain")),
  clip=F
))
ggsave(filename = "adults--all-scenarios-tall.png", width=9, height=8, plot=g)
ggsave(filename = "adults--all-scenarios-wide.png", width=9, height=6, plot=g)
```



